Application of Machine Learning to Increase the Efficiency of the Global Search Algorithm for Solving Multicriterial Problems\thanks{This research was supported by the Russian Science Foundation, project No. 21-11-00204.}

### 2023-07-07 2023/NUMTA/Abstract/Grishagin.tex

\institute{Lobachevsky State University of Nizhni Novgorod, Nizhni Novgorod, Russia \\
\email{\{konstantin.barkalov,evgeny.kozinov\}@itmm.unn.ru}, vagris@unn.ru}

@@

Decision making models described as problems of multicriterial optimization are very complicated for investigation because of the property of criteria contradictoriness that leads to the notion of the solution as the set non-dominated parameters (Pareto set). The complexity of these problems increases significantly in the case where criteria are multiextremal. In the approaches based on convolution schemes reducing the multicriterial problem to a family of scalar (single criterion) optimization subproblems, these subproblems are multiextremal as well. This circumstance requires the use of global optimization methods to solve those.

In the paper a novel algorithm for multicriterial black-box optimization with multiextremal criteria are considered. This algorithm applies ideas of complexity reduction when the initial multicriterial problem are reduced to a set of scalar subproblems on the base of maximum convolution, then each scalar multidimensional subproblem is transformed to equivalent one-dimensional problem by means of Peano mapping which provides single-valued correspondence between N-dimensional hyperparallelepiped and the unit interval of the real axis. For solving univariate problems a global optimization algorithm with guaranteed convergence to global optimum is used. As a core novel feature, this algorithm includes in its computational scheme a machine learning procedure with combination of accumulating the information of solved subproblems that allows one to significantly accelerate the building the Pareto set.

@@

\textbf{Acknowledgements.}

This research was supported by the Russian Science Foundation, project No. 21-11-00204.

>>>>

\institute{Lobachevsky State University of Nizhni Novgorod, Nizhni Novgorod, Russia \\
\email{\{konstantin.barkalov,evgeny.kozinov\}@itmm.unn.ru, vagris@unn.ru}}

@@

Decision making models described as problems of multicriterial optimization are very complicated for investigation because of the property of criteria contradictoriness that leads to the notion of the solution as the set non-dominated parameters (Pareto set). The complexity of these problems increases significantly in the case where criteria are multiextremal. 
%In the approaches based on convolution schemes reducing the multicriterial problem to a family of scalar (single criterion) optimization subproblems, these subproblems are multiextremal as well. This circumstance requires the use of global optimization methods to solve those.

In the paper a novel algorithm for multicriterial black-box optimization with multiextremal criteria are considered. This algorithm applies ideas of complexity reduction when the initial multicriterial problem are reduced to a set of univariate scalar subproblems on the base of Peano mapping and maximum convolution \cite{Gergel2018}.
%, then each scalar multidimensional subproblem is transformed to equivalent one-dimensional problem by means of Peano mapping which provides single-valued correspondence between $N$-dimensional hyperparallelepiped and the unit interval of the real axis. 
For solving univariate problems a global optimization algorithm with guaranteed convergence to global optimum is used \cite{Sergeyev2013}. As a core novel feature, this algorithm includes in its computational scheme a machine learning procedure with combination of accumulating the information of solved subproblems that allows one to significantly accelerate the building the Pareto set.

@@

\textbf{Acknowledgements.} 
This research was supported by the Russian Science Foundation, project No. 21-11-00204.

### 2023-07-07 2023/NUMTA/Full paper/NUMTA_2023_MCO.tex

All the competing methods solved the set of 100 MMCO problems several times such that in each case they executed different number of trials and, correspondingly, obtained different values of HV's. If we take a method, consider all the pairs $(K, HV(K))$ where $K$ is number of trials executed by the method and $HV(K)$ is the corresponding value of the HV-index, and place all these pairs as points on the plane with coordinate abscissa axis $K$ and ordinate axis $HV$ we build so called \textit{operational characteristic} \cite{Grishagin2016_2} of the method. On the base of operational characteristics it is easy to compare visually the efficiency of different methods. Indeed, if the operational characteristic of one method is located above the characteristic of the other, then the first method is better because it spends fewer trials for obtaining the same HV, or, equivalently, the first method provides better HV for the same number of trials. Operational characteristics of the competing algorithms for dimensions 3 and 4 are presented in Fig. \ref{fig:4} and \ref{fig:5} correspondingly.

>>>>

All the competing methods solved the set of 100 MMCO problems several times such that in each case they executed different number of trials and, correspondingly, obtained different values of HV's. If we take a method, consider all the pairs $(K, HV(K))$ where $K$ is number of trials executed by the method and $HV(K)$ is the corresponding value of the HV-index, and place all these pairs as points on the plane with the coordinate abscissa axis $K$ and the ordinate axis $HV$ we build so called \textit{operational characteristic} \cite{Grishagin2016_2} of the method. On the base of operational characteristics it is easy to compare visually the efficiency of different methods. Indeed, if the operational characteristic of one method is located above the characteristic of the other, then the first method is better because it spends fewer trials for obtaining the same HV, or, equivalently, the first method provides better HV for the same number of trials. Operational characteristics of the competing algorithms for dimensions 3 and 4 are presented in Fig. \ref{fig:4} and \ref{fig:5} correspondingly.
